Data Exploration: IUCN and AquaMaps

### 2015-12-19 data_explore.Rmd
update manuscript and graphs

@@

### define 25% and 75% quartiles for both x and y axes
area_ratio_quartiles <- quantile(spp_list$area_ratio, na.rm = TRUE)[2:4]
pct_overlap_quartiles <- quantile(spp_list$sm_perc, na.rm = TRUE)[2:4]

@@

### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, y = sm_perc, 
                                           color = popn_category,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point() + 
  scale_colour_manual(values = c('LC' = 'green4', 
                                 'NT' = 'greenyellow', 
                                 'VU' = 'yellow', 
                                 'EN' = 'orange', 
                                 'CR' = 'orangered2',
                                 'DD' = 'grey30')) +
  geom_vline(x = area_ratio_quartiles[c(1, 3)],
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles[c(1, 3)],
             color = 'grey50', linetype = 'dashed') +
#   geom_point(data = spp_list_label, shape = 4, color = 'grey30', size = 4) + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5) +
  labs(x = 'Area match: \n(range area smaller : range area larger) * 100%', 
       y = 'Distribution match: \n % overlap of smaller range w/ larger', 
       title = 'Area match vs distribution match',
       color = 'Risk', size = '% ocean range')

@@

```{r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

### Plot the bar chart
barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group_text = reorder(spp_group_text, pct_q1)), 
                                aes(x = spp_group_text, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin', alpha = .5) +
  scale_fill_manual(values = c('q1 well-matched'   = 'green4', 
                               'q2 dist-matched'   = 'greenyellow', 
                               'q3 area-matched'   = 'orange', 
                               'q4 poorly matched' = 'orangered2')) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0, size = 4) +
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Relative number of species per quadrant', 
       title = 'Species group breakdown by quadrant',
       fill = 'Match quadrant')

@@

### Plot the bar chart
barchart_spp_risk_quads <- ggplot(spp_risk_quadrants %>% 
                                  transform(popn_category = reorder(popn_category, cat_rank)), 
                                aes(x = popn_category, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin', alpha = .5) +
  scale_fill_manual(values = c('q1 well-matched'   = 'green4', 
                               'q2 dist-matched'   = 'greenyellow', 
                               'q3 area-matched'   = 'orange', 
                               'q4 poorly matched' = 'orangered2')) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0, size = 4) +
  coord_flip() +
  labs(x = 'Risk category', 
       y = 'Relative number of species per quadrant', 
       title = 'Risk category breakdown by quadrant',
       fill = 'Match quadrant')

print(barchart_spp_risk_quads)
ggsave(plot = barchart_spp_risk_quads,
       filename = file.path(dir_fig, 'barchart_spp_risk_quads.png'),
       width = 6, height = 4)

@@

iucn_maps_class <- spp_iucn_maps %>% 
  left_join(sciname_lookup %>% 
              select(sciname, class) %>%
              unique(), 
            by = 'sciname')

@@

knitr::kable(taxa_df %>%
               select(class, AquaMaps = n_am_class_all, IUCN = n_iucn_class_all, Both = n_class_both))
```

@@

taxa_df_long1 <- taxa_df %>%
  select(class, AquaMaps = n_am_class_all, IUCN = n_iucn_class_all, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both)
bar_taxa1 <- ggplot(taxa_df_long1, aes(x = class, y = n_spp)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity') +
  facet_grid(dataset ~ ., scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
print(bar_taxa1)

taxa_df_long2 <- taxa_df %>%
  select(class, AquaMaps = n_am_class_solo, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = .05), hjust = 0, size = 4, angle = 90, hjust = 1) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  labs(x = 'Class', y = 'Percent of mapped species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa2)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), height = 4, width = 8)

taxa_df_long3 <- taxa_df %>%
  select(class, AquaMaps = n_am_class_solo_ass, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa3 <- ggplot(taxa_df_long3, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = .05), hjust = 0, size = 4, angle = 90, hjust = 1) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  labs(x = 'Class', y = 'Percent of mapped and assessed species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa3)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_assessed.png'), height = 4, width = 8)

>>>>

Quadrant map and quad breakdown by taxa and risk

@@

### define 25%, 50%, and 75% quartiles for both x and y axes
area_ratio_quartiles <- quantile(spp_list$area_ratio, na.rm = TRUE)[2:4]
pct_overlap_quartiles <- quantile(spp_list$sm_perc, na.rm = TRUE)[2:4]

@@

### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, y = sm_perc, 
                                           color = popn_category,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point(alpha = .6)

### Manage scales for color and size 
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  scale_colour_manual(values = c('LC' = 'green4', 
                                 'NT' = 'greenyellow', 
                                 'VU' = 'yellow', 
                                 'EN' = 'orange', 
                                 'CR' = 'orangered2',
                                 'DD' = 'grey30')) + 
  scale_size_continuous(limits = c(0, 100),
                        breaks = c( 0,    25,    50,    75,    100), 
                        labels = c('0%', '25%', '50%', '75%', '100%')) + 
  scale_x_continuous(expand = c(0, 0), limits = c(0, 100)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100))

### here are median labels:
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  annotate(geom = 'text',
           x = area_ratio_quartiles[2], y = 5,
           hjust = 0, vjust = 0,
           color = 'grey50', size = 2,
           fontface = 'bold.italic', angle = 90,
           label = sprintf('Median = %s%%', round(area_ratio_quartiles[2], 1))) +
  annotate(geom = 'text',
           x = 5, y = pct_overlap_quartiles[2],
           hjust = 0, vjust = 0,
           color = 'grey50', size = 2,
           fontface = 'bold.italic', angle = 0,
           label = sprintf('Median = %s%%', round(pct_overlap_quartiles[2], 1)))

### here are median lines:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_vline(x = area_ratio_quartiles[2],
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = pct_overlap_quartiles[2],
#              color = 'grey50', linetype = 'dashed')

### here are 25% and 75% quantile lines:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_vline(x = area_ratio_quartiles[c(1, 3)],
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = pct_overlap_quartiles[c(1, 3)],
#              color = 'grey50', linetype = 'dashed')

### here are individual species, identified and named on the plot:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_point(data = spp_list_label, shape = 4, color = 'grey30', size = 4) + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5)

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  labs(x = bquote('Area alignment '*alpha['area']*' (%)'), 
       y = bquote('Distribution alignment '*alpha['dist']*' (%)'), 
       title = 'Distribution alignment vs area alignment',
       color = 'Risk', size = '% ocean range')

@@

``` {r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

### Plot the bar chart
barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group_text = reorder(spp_group_text, pct_q1)), 
                                aes(x = spp_group_text, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin', alpha = .5) +
  scale_fill_manual(values = c('q1 well-matched'   = 'green4', 
                               'q2 dist-matched'   = 'greenyellow', 
                               'q3 area-matched'   = 'orange', 
                               'q4 poorly matched' = 'orangered2')) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0, size = 2, color = 'grey20') +
  coord_flip() +
  labs(x = 'Taxonomic Group', 
       y = 'Relative proportion of species per quadrant', 
       title = 'Taxonomic group by quadrant',
       fill = 'Alignment')

@@

### Plot the bar chart
barchart_spp_risk_quads <- ggplot(spp_risk_quadrants %>% 
                                  transform(popn_category = reorder(popn_category, cat_rank)), 
                                aes(x = popn_category, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin', alpha = .5) +
  scale_fill_manual(values = c('q1 well-matched'   = 'green4', 
                               'q2 dist-matched'   = 'greenyellow', 
                               'q3 area-matched'   = 'orange', 
                               'q4 poorly matched' = 'orangered2')) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0, size = 2, color = 'grey20') +
  coord_flip() +
  labs(x = 'Risk category', 
       y = 'Relative proportion of species per quadrant', 
       title = 'Risk category by quadrant',
       fill = 'Alignment')

print(barchart_spp_risk_quads)
ggsave(plot = barchart_spp_risk_quads,
       filename = file.path(dir_fig, 'barchart_spp_risk_quads.png'),
       width = 6, height = 1.5)

@@

iucn_maps_class <- spp_iucn_maps %>% 
  left_join(sciname_lookup %>% 
              dplyr::select(sciname, class) %>%
              unique(), 
            by = 'sciname')

@@

knitr::kable(taxa_df %>%
               dplyr::select(class, 
                             AquaMaps = n_am_class_all, 
                             IUCN = n_iucn_class_all, 
                             Both = n_class_both))
```

@@

taxa_df_long1 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_all, IUCN = n_iucn_class_all, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both)
bar_taxa1 <- ggplot(taxa_df_long1, aes(x = class, y = n_spp)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity') +
  facet_grid(dataset ~ ., scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
print(bar_taxa1)

taxa_df_long2 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_solo, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 3), 
            size = 3, angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100))
  labs(x = 'Class', y = 'Percent of mapped species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa2)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), height = 4, width = 8)

taxa_df_long3 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_solo_ass, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa3 <- ggplot(taxa_df_long3, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 3), 
            size = 3, angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100))
  labs(x = 'Class', y = 'Percent of mapped and assessed species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa3)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_assessed.png'), height = 4, width = 8)
